Secure Composition of Robust and Optimising Compilers

### 2023-07-08 popl/main.tex
Finish up RelWork

  \item %
        % \Cref{sec:casestud} sets up a case-study to demonstrate the power of our formal framework.
        This paper presents a case-study showcasing the conjunction of the previous contributions (\Cref{sec:casestud}).
        %We define several different languages and compilers between them and prove that the compilers are robustly preserving different properties.
        To this end, it presents a compilation chain consisting of several passes.
        %The sequential composition of these compilers yields a compilation chain that is optimizing and robustly preserving a combination of \gls{ms} and \gls{cct}.
        The chain ultimately preserves a combination of \gls{ms} and \gls{cct} by means of composing the individual, secure passes concerning \gls{tms}, \gls{sms}, and \gls{scct}, respectively.
        %As for optimizations, the chain has two passes, one performing \gls{cf} and the other \gls{dce}.
        Furthermore, the chain includes two passes for \gls{ms}\MKin{maybe patch \gls{ms} when optimizing passes are done} that are optimizing: One performs a simple \gls{dce} and the other \gls{cf}.
        %This compilation chain demonstrates the ability to perform modular proof engineering by leaveraging this paper's compositionality theorems.
        This demonstrates that the usage of the presented framework is both simpler and more expressive compared to one monolithic security proof of the whole compilation chain.
        \MP{ matthis please write your idea here }

@@

Notably, real-world compilation chains also perform a series of passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimize the code or enforce a certain property.
Both examples can be seen in the wild, there are countless instances of them.\MK{cit}
Consider the compilation passes of LLVM\MK{cit} whose source-code optimisations are industry--competitive.
In LLVM, the pass manager can be used to schedule passes in a way where the optimisations yield the best results.
Take \gls{cf} and \gls{dce}.
Unfortunately, the order in which \gls{cf} and \gls{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration.
However, this costs additional time until compilation is done, which can be a valuable resource\MK{cit}.
Hence, compiler engineers typically try to find an order of optimizations that yields well-optimized programs as fast as possible.
The following corollary justifies that any such order of compilation passes is valid with respect to security.
So, given two compilation passes \bul{$\cc[_{1}]{\trg{L}}{\trg{L}}$, $\cc[_{2}]{\trg{L}}{\trg{L}}$, both robustly preserving class $\class[_{1}]$ or $\class[_{2}]$, respectively}, \oul{for any order of their composition the composed compiler robustly preserves the intersection of $\class[_{1}]$ and $\class[_{2}]$}.

@@

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls{dce}} (progoptcfdce);
    \end{tikzpicture}
  \end{center}
  \caption{Example program where the level of optimizations differ for one pass of applying \gls{cf} and \gls{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls{cf} or \gls{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimized towards the righthandside of the figure.}\label{fig:cfdceex}
\end{figure}

@@

      \draw[dashed] ($(sectms)-(0.8,0)$) -- ($(sectms)-(0.8,1.25)$);
      \draw[dashed] ($(sectms)-(-0.9,0)$) -- ($(sectms)-(-0.9,1.25)$);
      \draw[dashed] ($(secsms)-(-1.2,0)$) -- ($(secsms)-(-1.2,1.25)$);
      \draw[dashed] ($(secscct)-(1.2,0)$) -- ($(secscct)-(1.2,1.25)$);
      \draw[dashed] ($(secscct)-(-1.1,0)$) -- ($(secscct)-(-1.1,1.25)$);
    \end{tikzpicture}
  \end{center}
  \caption{Visualisation of the optimizing compilation pipeline that attains a combination of \gls{ms} and \gls{cct}.}\label{fig:pipeline}
\end{figure}
This section defines several secure compilers, each of which robustly preserves a different property of interest (\Cref{fig:pipeline}).
It demonstrates the power of the framework (\Cref{sec:compprop,sec:compcomp}) by composing these compilers for a secure and optimizing compilation chain that robustly preserves \gls{msscct}.
The first step in this chain is the compiler from $\src{L_{\tmssafe}}$ to $\trg{L}$ that robustly preserves just \gls{tms} (\Cref{thm:cca:rtp:tms}).
From here, an instrumentation from $\trg{L}$ to $\irl{L_{\mssafe}}$ ensures that no out-of-bounds accesses can happen and, thus, programs at this point attain \gls{sms} (\Cref{thm:ccb:rtp:sms}).
This is enough to get a compiler that robustly preserves \gls{ms} (\Cref{thm:ccab:rtp:ms}).
At this stage, the section presents two optimizing translations, namely \gls{cf} and \gls{dce}, each of which robustly preserves \gls{ms} (\Cref{thm:ccdce:rtp:ms,thm:cccf:rtp:ms}).
These translations can be freely ordered in the compilation chain without compromising memory safety (\Cref{thm:cccfccdce:rtp:ms}).
The last step ensures that code stays \gls{scct} (\Cref{thm:ccscct:rtp:scct}) which yields the final result that the whole compilation chain robustly preserves \gls{msscct} (\Cref{thm:ccall:rtp:msscct}).

@@

\subsection{Optimizing Compilers}\label{subsec:cs:opts}

@@

\MKin{category theory stuff}
\MKin{full abstraction -> not compositional}
\MKin{compositional compiler correctness?}

@@

This paper does a first step towards practical secure compilation chains by introducing a theoretical framework demonstrating that secure compilers compose.
The paper exercises this framework on a case-study that consists of an optimizing compilation chain that is secure with respect to \gls{msscct}.
In future work, it would be interesting to investigate wether it is possible to provide {\em secure compiler combinators}, similarily to how parser combinators work.
%To this end, it may be necessary to extend existing frameworks for multi--language semantics.
It is also interesting to extend our case--study and verify all results of it in Coq.
This way, it is possible to obtain an executable, secure compiler.
However, the formalization effort for backtranslation proofs is known to be enormous, so another research avenue may be to find better ways to (semi-)automate standard secure compilation proofs.

>>>>

  \item %
        % \Cref{sec:casestud} sets up a case-study to demonstrate the power of our formal framework.
        This paper presents a case-study showcasing the conjunction of the previous contributions (\Cref{sec:casestud}).
        %We define several different languages and compilers between them and prove that the compilers are robustly preserving different properties.
        To this end, it presents a compilation chain consisting of several passes.
        %The sequential composition of these compilers yields a compilation chain that is optimising and robustly preserving a combination of \gls{ms} and \gls{cct}.
        The chain ultimately preserves a combination of \gls{ms} and \gls{cct} by means of composing the individual, secure passes concerning \gls{tms}, \gls{sms}, and \gls{scct}, respectively.
        %As for optimisations, the chain has two passes, one performing \gls{cf} and the other \gls{dce}.
        Furthermore, the chain includes two passes for \gls{ms}\MKin{maybe patch \gls{ms} when optimising passes are done} that are optimising: One performs a simple \gls{dce} and the other \gls{cf}.
        %This compilation chain demonstrates the ability to perform modular proof engineering by leaveraging this paper's compositionality theorems.
        This demonstrates that the usage of the presented framework is both simpler and more expressive compared to one monolithic security proof of the whole compilation chain.
        \MP{ matthis please write your idea here }

@@

Notably, real-world compilation chains also perform a series of passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimise the code or enforce a certain property.
Both examples can be seen in the wild, there are countless instances of them.\MK{cit}
Consider the compilation passes of LLVM\MK{cit} whose source-code optimisations are industry--competitive.
In LLVM, the pass manager can be used to schedule passes in a way where the optimisations yield the best results.
Take \gls{cf} and \gls{dce}.
Unfortunately, the order in which \gls{cf} and \gls{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration.
However, this costs additional time until compilation is done, which can be a valuable resource\MK{cit}.
Hence, compiler engineers typically try to find an order of optimisations that yields well-optimised programs as fast as possible.
The following corollary justifies that any such order of compilation passes is valid with respect to security.
So, given two compilation passes \bul{$\cc[_{1}]{\trg{L}}{\trg{L}}$, $\cc[_{2}]{\trg{L}}{\trg{L}}$, both robustly preserving class $\class[_{1}]$ or $\class[_{2}]$, respectively}, \oul{for any order of their composition the composed compiler robustly preserves the intersection of $\class[_{1}]$ and $\class[_{2}]$}.

@@

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls{dce}} (progoptcfdce);
    \end{tikzpicture}
  \end{center}
  \caption{Example program where the level of optimisations differ for one pass of applying \gls{cf} and \gls{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls{cf} or \gls{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimised towards the righthandside of the figure.}\label{fig:cfdceex}
\end{figure}

@@

      \draw[dashed] ($(sectms)-(0.8,0)$) -- ($(sectms)-(0.8,1.25)$);
      \draw[dashed] ($(sectms)-(-0.9,0)$) -- ($(sectms)-(-0.9,1.25)$);
      \draw[dashed] ($(secsms)-(-1.2,0)$) -- ($(secsms)-(-1.2,1.25)$);
      \draw[dashed] ($(secscct)-(1.2,0)$) -- ($(secscct)-(1.2,1.25)$);
      \draw[dashed] ($(secscct)-(-1.1,0)$) -- ($(secscct)-(-1.1,1.25)$);
    \end{tikzpicture}
  \end{center}
  \caption{Visualisation of the optimising compilation pipeline that attains a combination of \gls{ms} and \gls{cct}.}\label{fig:pipeline}
\end{figure}
This section defines several secure compilers, each of which robustly preserves a different property of interest (\Cref{fig:pipeline}).
It demonstrates the power of the framework (\Cref{sec:compprop,sec:compcomp}) by composing these compilers for a secure and optimising compilation chain that robustly preserves \gls{msscct}.
The first step in this chain is the compiler from $\src{L_{\tmssafe}}$ to $\trg{L}$ that robustly preserves just \gls{tms} (\Cref{thm:cca:rtp:tms}).
From here, an instrumentation from $\trg{L}$ to $\irl{L_{\mssafe}}$ ensures that no out-of-bounds accesses can happen and, thus, programs at this point attain \gls{sms} (\Cref{thm:ccb:rtp:sms}).
This is enough to get a compiler that robustly preserves \gls{ms} (\Cref{thm:ccab:rtp:ms}).
At this stage, the section presents two optimising translations, namely \gls{cf} and \gls{dce}, each of which robustly preserves \gls{ms} (\Cref{thm:ccdce:rtp:ms,thm:cccf:rtp:ms}).
These translations can be freely ordered in the compilation chain without compromising memory safety (\Cref{thm:cccfccdce:rtp:ms}).
The last step ensures that code stays \gls{scct} (\Cref{thm:ccscct:rtp:scct}) which yields the final result that the whole compilation chain robustly preserves \gls{msscct} (\Cref{thm:ccall:rtp:msscct}).

@@

\subsection{Optimising Compilers}\label{subsec:cs:opts}

@@

While this paper focuses on the robust preservation framework~\cite{abate2019jour}, other secure compilation criteria exist.
The survey on formal approaches to secure compilation~\cite{patrignani2019survey} discusses a broad spectrum already, while this section presents a very high--level overview.
The full abstraction approach~\cite{abadi1999fullabstraction} states that a compiler should preserve and reflect observational equivalence between source and target programs.
It was shown~\cite{abate2021faandrc} that fully abstract compilers robustly preserve program properties that are either trivial or meaningless.
As a mitigation for this, the authors presented a categorical approach based on maps of distributive laws~\cite{watanabe2002modl}, which they call many maps of distributive laws.
Maps of distributive laws have been investigated before as a possible secure compilation criterion~\cite{tsampas2020catsc}.
Other approaches are extensions of the compiler correctness criterion as discussed in other work~\cite{patterson2019next700} or the introduction of opaque observations~\cite{vu2021reconciling} to reconcile compiler optimisations with security.
Note that this work also presents secure compilers that are optimising, but contrary to the other~\cite{vu2021reconciling}, provides a formal account of these in the robust preservation framework.

@@

This paper does a first step towards practical secure compilation chains by introducing a theoretical framework demonstrating that secure compilers compose.
The paper exercises this framework on a case-study that consists of an optimising compilation chain that is secure with respect to \gls{msscct}.
In future work, it would be interesting to investigate wether it is possible to provide {\em secure compiler combinators}, similarily to how parser combinators work.
%To this end, it may be necessary to extend existing frameworks for multi--language semantics.
It is also interesting to extend our case--study and verify all results of it in Coq.
This way, it is possible to obtain an executable, secure compiler.
However, the formalization effort for backtranslation proofs is known to be enormous, so another research avenue may be to find better ways to (semi-)automate standard secure compilation proofs.
